---
title: "FAQ & Troubleshooting"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{FAQ & Troubleshooting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Frequently Asked Questions

---

### General

**Q: What IFCB data formats are supported?**

A: The app reads standard IFCB files:

- ROI files (.roi) - Image data
- ADC files (.adc) - Metadata (dimensions, timestamps)
- HDR files (.hdr) - Header information

**Q: Can I use this for non-IFCB data?**

A: The app is specifically designed for IFCB data format. For other imaging systems, you would need to adapt the data loading functions.

**Q: Is my data modified?**

A: No. The app only reads your original files. All output is written to separate folders.

---

## Installation Issues

**Q: I see "Python not available" warning**

A: This warning affects saving .mat files. Python is required for:

- Saving annotations as .mat files for [ifcb-analysis](https://github.com/hsosik/ifcb-analysis)

Reading .mat files (annotations, classifier output, class lists) does not require Python. If you do not need to save .mat files, you can ignore this warning.

To enable .mat support:

```{r, eval = FALSE}
library(iRfcb)
ifcb_py_install()  # Creates venv in current working directory
```

Then restart the app.

**Q: Where is the Python virtual environment created?**

A: By default, `ifcb_py_install()` creates a `venv` folder in your current working directory. You can specify a different location:

```{r, eval = FALSE}
ifcb_py_install("/path/to/your/venv")
```

You can also specify the venv path when launching the app:

```{r, eval = FALSE}
run_app(venv_path = "/path/to/your/venv")
```

**Q: How is the Python virtual environment path resolved?**

A: The app uses the following priority order:

1. **`venv_path` argument** passed to `run_app()` (highest priority)
2. **Saved settings** from a previous session (stored in `settings.json`)
3. **Default** `./venv` in the working directory

When you specify `run_app(venv_path = "/path/to/venv")`, that path is used for Python initialization and pre-filled in the Settings dialog, overriding any previously saved path.

**Q: Package installation fails**

A: Make sure you have remotes installed and try:

```{r, eval = FALSE}
install.packages("remotes")
remotes::install_github("EuropeanIFCBGroup/ClassiPyR")
```

**Q: "Could not find app directory" error**

A: Try reinstalling the package:

```{r, eval = FALSE}
remotes::install_github("EuropeanIFCBGroup/ClassiPyR", force = TRUE)
```

**Q: iRfcb won't install**

A: [iRfcb](https://github.com/EuropeanIFCBGroup/iRfcb) is the core dependency for `ClassiPyR` and is installed automatically. If you encounter issues:

```{r, eval = FALSE}
remotes::install_github("EuropeanIFCBGroup/iRfcb")
```

---

## Data Loading

**Q: No samples appear in the dropdown**

A: Check that:

1. ROI Data Folder points to your data
2. Data is organized as: `folder/YYYY/DYYYYMMDD/files`
3. ROI files exist and are readable

**Q: "ROI file not found" error**

A: The app expects this structure:

```
roi_folder/
  2023/
    D20230101/
      D20230101T120000_IFCB134.roi
      D20230101T120000_IFCB134.adc
```

**Q: Classifications not loading**

A: For CSV files:

- Must have columns named `file_name` and `class_name` (exact names required)
- Optionally include a `score` column (confidence value between 0 and 1)
- The CSV file must be named after the sample (e.g., `D20230101T120000_IFCB134.csv`)
- File should be in the Classification Folder (searched recursively)

For MAT files:

- Must match pattern `*_class*.mat`
- Must contain `roinum` and `TBclass` variables
- Must contain `roinum` and `TBclass` variables

---

## CSV Format

**Q: What should my classification CSV look like?**

A: The CSV must have columns named `file_name` and `class_name`. The file must be named after the sample (e.g., `D20230101T120000_IFCB134.csv`).

Minimal example:

```
file_name,class_name
D20230101T120000_IFCB134_00001.png,Diatom
D20230101T120000_IFCB134_00002.png,Ciliate
```

With optional `score` column (confidence values between 0 and 1):

```
file_name,class_name,score
D20230101T120000_IFCB134_00001.png,Diatom,0.95
D20230101T120000_IFCB134_00002.png,Ciliate,0.87
D20230101T120000_IFCB134_00003.png,Dinoflagellate,0.72
```

**Q: My CNN classifier outputs different column names**

A: The column names must be exactly `file_name` and `class_name`. If your classifier uses different names, rename the columns before loading. For example in R:

```{r, eval = FALSE}
df <- read.csv("my_classifications.csv")
names(df)[names(df) == "predicted_class"] <- "class_name"
names(df)[names(df) == "filename"] <- "file_name"
write.csv(df, "D20230101T120000_IFCB134.csv", row.names = FALSE)
```

---

## Annotation

**Q: Drag-select isn't working**

A: Make sure you're:

1. Starting the drag in the gallery area (not on an image)
2. Using left mouse button
3. Dragging far enough (>5 pixels)

**Q: Images show as "Not found"**

A: The ROI might be empty (no actual image data). These are filtered out automatically on load.

**Q: Changes aren't being saved**

A: Check that:

1. Output folder is writable
2. Python is available (required for saving .mat files)
3. Click "Save Annotations" before closing

---

## Class List

**Q: Do I need to upload a class list file to start annotating?**

A: No! You can create a class list from scratch directly in the app:

1. Open Settings → Edit Class List (no file upload needed)
2. Add classes one at a time or paste multiple classes in the text area
3. Click Apply Changes - a temporary file is created automatically
4. Start annotating immediately
5. Remember to Save as .mat or .txt for future sessions

**Q: How do I create a class list from scratch?**

A:

1. Open Settings → Edit Class List
2. Add classes using "Add new class" field, or type/paste classes in the text area
3. Click Apply Changes
4. Save as .mat or .txt for future use

**Q: Can I import a class list from MATLAB?**

A: Yes, load your existing `class2use.mat` file via Settings.

**Q: My class names look different after loading**

A: The app may truncate trailing numbers from class names (e.g., "Diatom_01" → "Diatom"). This matches iRfcb behavior.

**Q: Why are some characters in my class names changed?**

A: Class names are used as folder names when exporting PNGs, so certain characters must be sanitized:

| Character | Reason | Action |
|-----------|--------|--------|
| `/` or `\` | Path separators | Replaced with `_` |
| `< > " ' &` | HTML/security risks | Removed |
| `: * ? \|` | Windows filesystem | Removed |
| `..` | Path traversal risk | Removed |

**Example**: `Snowella/Woronichinia` becomes `Snowella_Woronichinia`

Common taxonomic characters like hyphens (`-`), underscores (`_`), periods (`.`), and spaces are preserved.

---

## Output

**Q: Where are my annotations saved?**

A: In the Output Folder you configured:

- MAT annotation files are saved directly in the output folder (one per sample)
- `validation_statistics/` subfolder contains CSV statistics
- PNGs are in the PNG Output Folder, organized by class name

**Q: Can I import annotations back to MATLAB?**

A: Yes, the MAT files are compatible with the [ifcb-analysis](https://github.com/hsosik/ifcb-analysis) toolbox (Sosik & Olson, 2007). Use:

```matlab
load('sample_name.mat');
% classlist contains [roi_number, class_index]
```

Note: Python with scipy must be installed to save .mat files.

**Q: What's in the statistics CSV?**

A: Two files per sample:

1. `*_validation_stats.csv` - Summary statistics
2. `*_validation_detailed.csv` - Per-image details

---

## Measurement

**Q: How do I measure images?**

A: Click the Measure button (ruler icon) in the toolbar to activate measure mode. Then click and drag on any image to draw a measurement line. The distance is displayed in micrometers and pixels.

**Q: How do I change the scale calibration?**

A: Open Settings and find "Pixels per Micrometer". The default is 3.4 (standard for IFCB). Enter your instrument's calibration value.

---

## Settings & Persistence

**Q: Are my settings saved between sessions?**

A: Yes! Settings are stored in a configuration file:

- **Linux**: `~/.local/share/ClassiPyR/settings.json`
- **macOS**: `~/Library/Application Support/ClassiPyR/settings.json`
- **Windows**: `%LOCALAPPDATA%/ClassiPyR/settings.json`

Folder paths, class list location, and Python venv path are automatically restored when you restart the app.

**Q: How do I reset all settings to defaults?**

A: Use the `reset_settings` argument when launching the app:

```{r, eval = FALSE}
run_app(reset_settings = TRUE)
```

This deletes the saved `settings.json` file and starts the app with default values. This is useful if:

- The app fails to start due to invalid saved paths
- Folder paths point to locations that no longer exist
- You want a clean slate after changing your data layout

You can also combine it with other arguments:

```{r, eval = FALSE}
# Reset settings and specify a new Python environment
run_app(reset_settings = TRUE, venv_path = "/path/to/your/venv")
```

**Q: What's the yellow warning on some classes?**

A: Classes marked with a warning are in your classification data but not in your class list. This can happen when:

- Loading classifications from a classifier trained with a different class list
- The class list has been modified since the classification was created

You can still work with these classes, but consider adding them to your class list for consistency.

**Q: A sample shows both annotation and classification markers - what does that mean?**

A: This sample has both manual annotations AND auto-classifications available. When you load it:

- The app defaults to manual annotation mode
- You can switch to validation mode using the button in the header
- Each mode maintains its state independently

---

## Error Messages

| Error | Solution |
|-------|----------|
| "ROI file not found" | Check ROI Data Folder path and file structure |
| "ADC file not found" | ADC file must be alongside ROI file |
| "Python not available" | Affects saving .mat files. Run `iRfcb::ifcb_py_install()` |
| "Error loading class list" | Check file format (.mat or .txt) |
| "No samples found" | Check ROI Data Folder configuration |
| App fails to start | Try `run_app(reset_settings = TRUE)` to clear saved settings |

---

## Performance Tips

1. **Use pagination** - Lower images per page for faster loading

2. **Filter by class** - Reduces rendering load

3. **Close other apps** - Image extraction uses memory

4. **SSD storage** - Faster file access

---

## Getting Help

- [GitHub Issues](https://github.com/EuropeanIFCBGroup/ClassiPyR/issues) - Report bugs
- [iRfcb Documentation](https://github.com/EuropeanIFCBGroup/iRfcb) - Core dependency for IFCB data handling
