---
title: "Annotating Images from Any Instrument (Generic Mode)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Annotating Images from Any Instrument (Generic Mode)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

ClassiPyR was originally built for the Imaging FlowCytobot (IFCB), but its
gallery-based, drag-select annotation workflow is useful for **any** image
labelling task. **Generic instrument mode** removes the IFCB-specific
assumptions so you can annotate a folder of ordinary image files — from a
microscope, a camera trap, a benthic survey, a histology scanner, or any other
source — and even non-scientific image-classification tasks.

This vignette explains how generic mode differs from IFCB mode, how to lay out
your images, and how annotations are stored.

---

## When to use generic mode

Use **generic mode** when:

- Your images are already exported as ordinary files (PNG, JPG, JPEG, ...).
- Your file and folder names do **not** follow the IFCB convention
  (`DYYYYMMDDTHHMMSS_IFCBNNN`).
- You do not need MATLAB (`.mat`) output for the `ifcb-analysis` toolbox.

Use the default **IFCB mode** when you work with raw IFCB data (`.roi`/`.adc`),
the IFCB Dashboard, or need `.mat` export. IFCB mode is unchanged and remains
the default — switching to generic mode does not affect IFCB workflows.

---

## Switching to generic mode

1. Open **Settings**.
2. Under **Instrument**, choose **Generic (any image source)**.
3. Optionally adjust **Image file extensions** (default: `png, jpg, jpeg`).
4. Set the **Image / ROI Data Folder** to the folder containing your images.
5. Click **Save Settings**. The app re-scans the folder automatically.

In generic mode the IFCB-specific controls (Dashboard, classification folder,
threshold, MAT export, live CNN prediction) are hidden, and annotations are
always stored in SQLite.

---

## Organising your images

A **sample** in ClassiPyR is simply a folder of images. Generic mode supports
two layouts.

### 1. Flat folders (annotate from scratch)

Point the data folder at a directory whose subfolders each contain a set of
images. Each subfolder becomes one sample, named after the folder:

```
my_images/
├── site_A_2024-05-01/
│   ├── img_0001.jpg
│   ├── img_0002.jpg
│   └── close up (2).jpg
└── site_B_2024-05-02/
    ├── scan01.png
    └── scan02.png
```

Here `site_A_2024-05-01` and `site_B_2024-05-02` are two samples. All images
start as `unclassified`, ready for you to label them in the gallery.

A single flat folder of images also works: point the data folder directly at
`site_A_2024-05-01/` and that folder itself is treated as one sample.

Any folder name is accepted as long as it is safe to use as a folder name (no
path separators, no `..`, and none of the characters `< > " ' & : * ? |`).

### 2. Class-organised folders (bulk import existing labels)

If your images are already sorted into class subfolders, you can bulk-import
them via **Settings → Import to SQLite → Images → SQLite**:

```
labelled_export/
├── Cat/
│   ├── photo_01.jpg
│   └── photo_02.jpg
└── Dog/
    └── rex.jpeg
```

Each top-level subfolder is treated as a **class**. Every image becomes its own
single-image sample (the sample name is the file name without its extension),
so each image carries its own label. This is convenient for reviewing or
correcting an existing classification.

---

## How images are identified

IFCB images are named `SampleName_NNNNN.png`, where `NNNNN` is the ROI number.
Generic images can have **any** name, so ClassiPyR assigns each image a stable
*ROI number* derived from the alphabetical order of the file names within its
folder.

Crucially, the real file name is stored in the annotations database, so your
arbitrary names (including names with spaces) round-trip exactly and your
labels stay attached to the correct image across reloads.

---

## Where annotations are stored

Generic mode always uses the **SQLite** backend (`annotations.sqlite` in the
Database Folder). This is portable, requires no Python, and is the same database
used by IFCB mode. You can query it directly, or use the package's loading
functions:

```{r eval=FALSE}
library(ClassiPyR)

db_path <- get_db_path("/path/to/db_folder")

# List samples that have annotations
list_annotated_samples_db(db_path)

# Read annotations for one sample (file names are returned as stored)
load_annotations_db(db_path, "site_A_2024-05-01", roi_dimensions = NULL)
```

The `annotations` table stores `sample_name`, `roi_number`, `class_name`,
`annotator`, `timestamp`, and `file_name`. Because the file name is stored,
annotations made in generic mode are fully self-describing.

---

## Programmatic import

You can import a class-organised folder without the app:

```{r eval=FALSE}
db_path <- get_db_path("/path/to/db_folder")

import_png_folder_to_db(
  png_folder = "/data/labelled_export",
  db_path = db_path,
  class2use = c("Cat", "Dog"),
  annotator = "reviewer",
  instrument_type = "generic",
  image_extensions = c("png", "jpg", "jpeg")
)
```

---

## Limitations in generic mode

- No `.mat` (MATLAB / `ifcb-analysis`) export — SQLite only.
- No IFCB Dashboard integration or live CNN prediction.
- The year/month/instrument filters (which rely on IFCB date-stamped names) are
  disabled; all samples are listed together.

These are intentional: they keep generic mode simple and instrument-agnostic
while leaving the full IFCB workflow untouched in IFCB mode.

### Sample folder names must be unique

A sample is identified by its folder name, not its full path. If two folders
anywhere under the image folder share the same name (for example
`2023/StationA` and `2024/StationA`), only the **first** one found is used as a
sample and the others are ignored — a warning lists the duplicates during the
scan. Give each image folder a unique name (e.g. `StationA_2023`,
`StationA_2024`) so every folder is annotatable.

### Adding images to an already-annotated sample

Annotations are matched to images by file name, so a sample is loaded from the
set of images that existed when it was last saved. If you add **new** image
files to a folder that already has saved annotations, those new images are not
picked up on reload (the previously stored annotations are shown instead). Add
all of a sample's images before annotating it; if you must extend a sample
later, remove its existing annotations first so the full folder is re-discovered,
or re-import the folder with `import_png_folder_to_db()`.

### Navigating large sample sets

Because generic mode does not assume a timestamp in your file or folder names,
the IFCB year → month → instrument drill-down is unavailable. Samples are
instead chosen from a single, alphabetically sorted dropdown (which is
searchable — start typing a sample name to filter it) together with the
annotated / not-yet-annotated status filter.

This is comfortable for tens or hundreds of samples, but with **thousands** of
samples the flat list can be harder to navigate than the IFCB date hierarchy.
If this affects your workflow, a practical workaround is to encode a grouping
key in your folder names (for example a date or site prefix such as
`2024-05-01_siteA`) so that related samples sort together and are easy to find
by typing in the dropdown.
